An Extensible Approach to Verification of Embedded Network Systems*

### 2014-04-20 tech-project/paper.tex
updating paper

% CpSc 855 - Final paper
% Takumi Bolte & Dan Welch - Spring 2014.
\documentclass{sig-alternate}
%\usepackage{flushend}		% For automatically balancing reference columns

\begin{document}
\conferenceinfo{Clemson 2014}{7th Clemson University Mini-Conference on Embedded Network Systems}
\title{An Extensible Approach to Verification of Embedded Network Systems*}

@@

In this paper we present a flexible means of specifying and verifying the correctness of software for embedded network systems. Our approach uses RESOLVE, an imperative, component based programming and mathematical specification language, to verify the functional correctness of TinyOS applications. In doing so, we enrich the work originally presented in \cite{regula:2010} with the following: A model view controller (MVC) based implementation of a RESOLVE to C translator, a dynamic memory allocation scheme tailored towards embedded systems running the code generated by our tool, and the addition of a new language keyword that enables users to pair custom RESOLVE specifications with `externally' implemented (non RESOLVE) realizations. We demonstrate these additions with a revised version of the LED template originally presented in \cite{regula:2010} that showcases some of the more recent mathematical developments in ``string theory" (resolve's mathematical string library) as well as verification of a simple enhancement for Leds template that includes preliminary generation of performance profile characteristics.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Within little more than a decade, the area of embedded network systems and wireless sensing has exploded in popularity within industry and academia alike. Tempering however this extremely quick rise in popularity is the inherent difficulties in developing applications that function as intended in low power, event-driven environments. In response to these difficulties, a variety of tools and languages have been put forth  to help ease the burden on developers. On one end of this effort are languages such as NesC (Network embedded sensor C) which strive to minimize concurrency issues and other common sources of error by hiding libraries of pre-written drivers underneath hierarchies of software and interface level abstractions. The other end is largely comprised of simulation tools such as TOSSIM, Cooja, and Arora that make use of high-fidelity simulations of mote network and hardware behavior to model networks offline in controlled, repeatable environments.

Though these and other tools have indeed proven invaluable in allowing users to test and reason about event-driven code prior to deployment, they remain incapable of providing complete assurance that code will behave as expected when deployed in the field.

In this paper, we approach this problem by using RESOLVE (Reusable Software Language with Verification) [CITE] [CITE] [CITE] as means of specifying, writing, and ultimately verifying proposed embedded code at a high level, then present a technique of translating this verified code down into lower level, embedded-platform friendly C.

Our choice to use RESOLVE as a language and verification frontend stems from the language's well acknowledged characteristics [CITE] that emphasize the following:
\begin{itemize}
\item \textbf{Modularity} At the language level, a strict separation of concerns is enforced between module specifications and client implementations. Thus, for any one particular specification, there can be any number of interchangeable implementations. This separation is ideal considering that many embedded applications are composed in similar fashion: Various drivers oftentimes provide a fixed set of operations, but have many distinct implementations that vary from platform to platform, vendor to vendor.

\item \textbf{Mathematical flexibility} RESOLVE offers a flexible mathematical typing system that allows users to either draw from a library of preexisting mathematical units when writing specifications, or create their own.

@@

This paper aims to leverage RESOLVE's mathematical flexibility of and component based design principles to the challenge of formally verifying an LED driver for the Telos motestack family.

We discuss this challenge in three parts. In the first, we provide both a short overview of the Telos family mote platform -- including a description of the LEDs driver which we later specify, verify, and translate -- as well as a short overview of the RESOLVE language itself. In the second section, we provide revised specifications for the LED drivers as well as a single enhancement on these drivers. This is followed by a review discussing the results of verification along with a review of some verification conditions generated by the compiler involving preliminary results for performance characteristics of enhancements. In the third and final section of the paper, we discuss the tool itself -- providing implementation details as well as a discussion of how this iteration of the tool differs from that originally detailed in \cite{regula:2010}. Finally, we conclude by discussing tool improvements and directions for future research.
\begin{figure*}
\centering
\includegraphics[scale=.55]{figs/ast_traversal.pdf}
\caption{An AST of a RESOLVE operation and its corresponding AST treewalker call sequence.}
\end{figure*}
\label{fig:ast}
\section{A model for c translation}

One of the primary challenges in translating from RESOLVE to C is finding a suitable C representation for each RESOLVE module and the constructs allowable in each. As we are dealing with an environment where functional correctness is a primary concern, it is especially important that the code generated represents as closely as possible the original RESOLVE source. In an effort to make such considerations, at the highest level, our tool outputs code specifically tailored to one of three distinct RESOLVE module types.

\begin{description}
\item [Concepts]
To model as closely as possible the role of a concept module, our tool produces a single .h file that contains structures for each parameter to a concept, pointers corresponding to the abstract types defined by the concept, as well as set of function pointers for each of the methods provided by the original RESOLVE concept.

\item [Realizations] 
Referring to both enhancement realizations as well as concept realizations, these are translated into a .h/.c pair. The header portion of this pair contains function declarations for the creation and destruction of a ``Realization\_Specific" facility. Elements of this facility are accessible only through a concept interface. The .c file has 5 sections:

\item [Facilities]

\end{description}

The representation of objects make use of a special

\section{Implementation}
Development of the tool presented here can be logically partitioned into two phases: The tool that performs the actual translation from RESOLVE to C, and a separate tool that is called by the generated code that is responsible for allocating appropriate amounts of memory on the motestack without needing to resort to static allocation.

\subsection{Translation AST walking}
Translation is performed over the course of a traversal of RESOLVE's abstract syntax tree (AST). The traversal mechanism used is a dervivative of the visitor pattern that provides a SAX-dom style pre-post traversal over all nodes in the tree. Thus, for any given node present, a total of two visits occur: One corresponding to the node being `hit' during the pre traversal stage, and one for the post.

@@

\begin{verbatim}
operation nothing() procedure;
    var x : Integer;
    x := 3;
end nothing;
\end{verbatim}

Shown in Figure \ref{fig:ast} is the AST representation of operation \texttt{nothing}. Here, language constructs are represented as labeled boxes, while the actual traversal over these constructs -- and the order in which it is performed -- is communicated on the right via the call stack. Each of these calls are received in the translator in the order in which they are visited within the tree. It is up to the client (in this case, the author of the C-translator) to decide which of these methods they wish to override and perform custom actions within.

We feel this particular traversal pattern lends itself well to the this application for the following reasons:

\begin{itemize}
\item A visit method for a construct provides a convenient encapsulation of all the logic required to translate RESOLVE construct $x$ to an appropriate C construct $y$.

@@

\item Finally, only the visit methods for the constructs we currently wish to process need to be overridden. This makes it especially easy to tweak and optimize the size of the codebase for the C translator.
\end{itemize}

\subsection{Translation output}
Output of translated code is done using \textit{Stringtemplate} -- a third-party tool written in Java that allows users to define parameterizable templates. Like the name suggests, a template is simply ``a document with holes" that the user choses when and how to fill.

An example of a template for a function/operation definition might look like the following:

@@

Words enclosed in \texttt{<..>} signifies an attribute passable to the template and indicates where it will be positioned. It is entirely up to the user to define which attributes they wish to fill in, and at what level of granularity they wish to compose their templates. For example, the user might opt to fill-in and pass the \texttt{parameter(..)} template shown below to the \texttt{params} attribute, or, alternatively, a simple string.

@@

In the context of language translation, these templates when stored on a stack and manipulated over the course of the AST traversal, simplify the task of producing arbitrarily complicated, nested blocks of C output. For instance, upon visiting \texttt{preOperationDec}, a \texttt{function\_def} template can be instantiated by the client and pushed onto a global translation stack with its \texttt{name}, return \texttt{type}, and \texttt{modifier} attributes filled in. As \texttt{preOperationDec}'s children get visited, the \texttt{function\_def} template currently at the top of the stack receives similarly constructed parameter, variable, and statement templates from the corresponding child nodes. Upon reaching \texttt{postOperationDec}, we can be assured that the function has been completely filled in -- assuming the user implemented the visit methods for the children.

Hence, the only work being done within visit methods themselves is forwarding appropriate information from tree-nodes to an externally defined template. This allows us to exploit (in shameless design pattern parlance) a strict model view controller (MVC) separation in our translator's codebase between the mechanism that does the AST visiting (controller), the tree nodes from which we're adding information to templates (model), and the external file containing all available C language templates (view).

@@

\subsection{Garbage collection}

a source to source translator which, given a resolve component, outputs a C representation of that component.

@@

Special thanks to Mike Kabanni and Mark Todd whose initial contributions and suggestions for the model of C translation paved the way to this paper.

>>>>

% CpSc 855 - Final paper
% Takumi Bolte & Dan Welch - Spring 2014.
\documentclass{sig-alternate}

\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{listings}
\begin{document}
\conferenceinfo{Clemson 2014}{7th Clemson University Mini-Conference on Embedded Network Systems}
\title{An Extensible Approach to Verification of Embedded Network Systems*}

@@

In this paper we present a flexible means of specifying and verifying the correctness of software for embedded network systems. Our approach uses RESOLVE, an imperative, component based programming and mathematical specification language, to verify the functional correctness of TinyOS applications. In doing so, we enrich the work originally presented in \cite{regula:2010} with the following: A model view controller (MVC) based implementation of a RESOLVE to C translator, a dynamic memory allocation scheme tailored towards embedded systems running the code generated by our tool, and the addition of a new language keyword that enables users to pair custom RESOLVE specifications with `externally' implemented (non-native RESOLVE) realizations. We demonstrate these additions on an LED (Light emitting diode) driver that showcases recent mathematical developments, as well as formal verification of two enhancements that we demonstrate running on a Telos mote.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
Within little more than a decade, the area of embedded network systems and wireless sensing has exploded in popularity within industry and academia alike. Tempering however this extremely quick rise in popularity is the inherent difficulties in developing applications that function as intended in low power, event-driven environments. In response to these difficulties, a variety of tools and languages have been put forth  to help ease the burden on developers. On one end of this effort are languages such as NesC, (Network embedded sensor C) which strive to minimize concurrency issues and other common sources of error by hiding libraries of pre-written drivers underneath hierarchies of software and interface level abstractions. The other end of this effort is largely comprised of simulation tools such as TOSSIM, Cooja, and Arora that make use of high-fidelity simulations to model networks offline in controlled, repeatable environments. Though these and other tools have indeed proven invaluable in allowing users to test and reason about event-driven code prior to deployment, they remain incapable of providing complete assurance that code will behave as expected when deployed in the field.

We approach this problem by using RESOLVE (Reusable SOftware Language with VErification) as means of authoring, specifying, and ultimately verifying code for embedded network systems. Our decision to use RESOLVE as a language frontend -- as opposed to verifying C code directly -- ultimately stems from a verification amenability standpoint: Not only does RESOLVE prohibit verification crippling operations such as uncontrolled referencing and aliasing (prevalent in C and many other current languages)\cite{kulczycki:2004}, but also embodies a number of other characteristics ideal for embedded platforms including:

\begin{itemize}
\item \textbf{Modularity} RESOLVE enforces a strict separation of concerns between module specifications and client implementations. As a result, for any one particular specification, there can be any number of interchangeable implementations. This separation is ideal considering that many embedded applications happen to fit this pattern nicely: Various drivers oftentimes provide a common set of functionality, but in general have many distinct implementations that vary arbitrarily from platform to platform, vendor to vendor.

\item \textbf{Mathematical flexibility} RESOLVE offers a rich, mathematical type system that allows users to either draw from a library of preexisting mathematical units when writing specifications, or simply create their own. This ideal in a setting where various driver applications are varied and new contexts might require creating new theory constituents appropriate for specific drivers. % reword this to be less vague.

@@

The paper is organized as follows: First, we open with a brief overview of the Telos mote platform. Next, we present revised specifications of an LED driver component with a formally verified enhancement. This section is concluded with a review of verification results, and a discussion of any relevant theorems and verification conditions (VCs) used. The remainder of the paper is spent detailing the model of C code generated by the tool, giving a quick overview of generation process itself, and detailing a dynamic, stack-based memory allocation tool utilized by the translated code. We conclude with suggestions for tool improvements, as well as a review of some longer term research goals.

\section{The Telos Platform}

The Telos mote \cite{polastre:2005} is a programmable, low power, wireless sensing device developed at UC Berkeley. Its hardware includes an msp430 microcontroller containing 128 bytes of RAM and 10kB of programmable flash memory. A cc2420 radio stack provides the mote with broadcast and receiving capabilities, while optional sensing capabilities may be added in the form of light, humidity, and temperature sensors. The mote also contains three onboard LEDs: One blue, one red, and one green.

%Additionally, the Telos board contains a total of three LEDs -- one red, one green, and one blue.
\section{Specifiying LED behavior in Resolve}
\label{sec:specifiying}

In RESOLVE, programs are composed of several different modules that range from interfaces and realizations, to client (facility) modules. In this section, we provide mathematical and programmatic elements of an LED strip component to give readers both a concrete look at the language, and introduce some new features aimed at making it more amenable to the development of embedded applications. Note that while we provide the level of detail necessary to understand the current example, readers interested in gaining a more complete, in depth knowledge of the language are encouraged to refer to \cite{sitaraman:2011, kulczycki:2008}.

\subsection{Concepts}

A \textit{concept} module in RESOLVE defines a specification for a mathematical, abstract type. Similar to an interface in Java, a concept provides a number of operation signatures that implementors are expected to realize. Shown below is a \texttt{LED\_Template} concept that provides a light strip abstraction.

\begin{verbatim}
Concept LED_Template(eval Strip_Length: Integer);
    uses Boolean_Theory, String_Theory;
    requires 0 < Strip_Length <= 4;

    Type Family LED is modeled by Str(B);
        exemplar L;
        constraint |L| = Strip_Length;
        initialization ensures 
            ...

    Oper Set(updates L : LED; eval b : Boolean; 
                             eval i : Integer);
        requires 0 <= i < |L|;
        ensures |L| = |#L| and 
                   Element_At(i, L) = b;

    Oper Status(preserves L : LED; 
                   eval i : Integer) : Boolean;
        requires 0 <= i < |L|;
        ensures Status = Element_At(i, L);

end LEDs_Template;
\end{verbatim}

We model our conceptual LED strip using a mathematical string (finite sequence) of booleans, denoted \texttt{Str(B)}, where each boolean within the string indicates the status of that particular LED: On (true) or off (false). The \texttt{exemplar} clause located immediately below provides a handle to this abstract model, and is used throughout the remainder of the specification.

%The \texttt{Type Family} clause introduces an abstract type for the LED strip which we mathematically model using strings of booleans, denoted \texttt{Str(B)}, while the \texttt{exemplar} declared immediately after provides a handle to this type family. Use of the \texttt{exemplar} can be observed in the \texttt{constraint} clause, where it is used to assert that length of the strip is fixed to be whatever the user has specified via the \texttt{Strip\_Length} parameter. Finally, the \texttt{initialization} \texttt{ensures} clause makes explicit the state of the structure at the time it is initialized. We omit this verification specific clause, as it has little bearing on verification due to the fact that the realization we provide is \textit{externally} realized -- a point discussed in the next section\footnote{This clause, if it were provided, would need to communicate the following: ``For each position $i$ within the strip, the LED (or, value) at position $i$ is initially false."}

It's worth noting that unlike the \texttt{LED\_Template} presented in \cite{regula:2010} which models an LED as the \textit{cartesian product} of booleans $b_1$, $b_2$, \ldots , $b_4$, the strip model we present here instead uses strings for the following reasons:
\begin{itemize}
\item Strings are indexible, and thus do not require separate \texttt{Set} and \texttt{Status} operations for each individual LED.
\item This approach demonstrates the benefits of reusable mathematical theories. The specifications here are based solely in \texttt{String\_Theory}, a pre-existing RESOLVE math library that provides all the definitions and theorems used in this specification.
\end{itemize}

The concept provides two operations. The first of the two, \texttt{Set}, takes as a parameter an instance of an LED strip \texttt{L}, a boolean \texttt{b}, and an integer \texttt{i}. The operation \texttt{requires} that \texttt{i} falls within the length of the strip, and \texttt{ensures} upon completion that:
\begin{enumerate}
\item The length of the outgoing strip \texttt{L}, is the same length as the incoming strip, \texttt{\#L} (where \texttt{\#} signifies its incoming state).
\item The LED in position \texttt{i} of \texttt{L} is set to boolean parameter \texttt{b}. 
\end{enumerate}
Operation \texttt{Status}, which simply returns the state of the LED at position \texttt{i}, is specified similarly.

%Each operation carries with it a pre and post condition in the form of \texttt{requires} and \texttt{ensures} clause, which make explicit what must be true coming into the function, and what must be true after. For instance, the requires clause to \texttt{Status} states that \texttt{i} must fall somewhere within the bounds of the string which models our LED strip, while the \texttt{ensures} clause makes use of a \texttt{String\_Theory} definition, \texttt{Element\_At}, to assert that the boolean returned from \texttt{Status} is indeed the value occupying position \texttt{i} of \texttt{L}. It's important to note that the variables appearing in the \texttt{requires} and \texttt{ensures} clauses are strictly mathematical values. Thus, the \texttt{L} parameter of operation \texttt{Status} is referring to the mathematical value of an \texttt{LED}, as opposed to a programmatic one (typically defined in a realization).

%\subsection{An External LED Realization}

%Armed with a specification, the RESOLVE programmer would now ideally provide a verifiable implementation of \texttt{LEDs\_Template} via a RESOLVE realization. However, as our target platform is embedded, and our concept  aims to provide control for LEDs -- a decidedly low level feature -- our realization is forced to operate at similarly low levels by manually interfacing with hardware pins provided by msp430 chipset. RESOLVE, however in its current state is far too high level of a language to perform these tasks directly -- that is, it currently lacks driver support to do so. To address this difficulty, we introduce language support for \textit{external realizations}, which, in essence, leaves the user free to write their own realization of a concept in whatever language is most convenient. In the case of \texttt{LEDs\_Template}, we provide a hand written \texttt{Telos\_LEDs\_Realiz} C realization that matches as closely as possible the model of translation detailed in Section \#.

%More than givingHaving the freedom to use RESOLVE's specificational capabilities

\subsection{Enhancements}

RESOLVE also allows users to extend the functionality provided by the base concept through \textit{enhancements} -- a form of specification inheritance. The enhancement we provide here is  \texttt{Toggling\_Capability}, which gives users of the LED component the ability to toggle a specific LED on or off.

Shown below is a specification for \texttt{Toggling\_Capability} and one particular realization of it.

\begin{verbatim}
Enhancement Toggling_Capability for LEDs_Template;

    Oper Toggle(upd L : LED; eval i : Integer);
        requires 0 <= i < |L|;
        ensures Element_At(i, L) = 
                        not(Element_At(i, #L));
end Toggling_Capability;

Realization Toggling_Realiz for
            Toggling_Capability of LEDs_Template;

    Proc Toggle(upd L : LED; eval i : Integer);
        Var Content : Boolean;

        Content := Status(L, Replica(i));
        Set(L, Not(Content), Replica(i));
    end Toggle;

end Toggling_Realiz;
\end{verbatim}

The specification of \texttt{Toggle} states that upon termination, the LED located at position \texttt{i} in \texttt{L} is the complement of that same location in the incoming LED, \texttt{\#L}.

Note that the enhancement specifications themselves look and function largely the same as a normal concept: Each specifies a purely conceptual module, and hence is implementation neutral.

Even realizations of enhancements are neutral since they are only permitted to call operations provided by the base concept -- thus requiring no knowledge of how the called operation is actually implemented.

\subsection{Verification}

In this section we verify our realization of \texttt{Toggling\_Capability}. The first step in this process is to generate verification conditions (VCs) for \texttt{Toggling\_Realiz}, which, if proven, will establish the correctness of that particular realization\footnote{Note: VCs themselves are typically generated from specific lines of a realization to ensure that the overall content of a realization is consistent with its specification.}.

\begin{figure}[!htb]
\centering
\begin{tabular}{lccc}
	\toprule
	Condition \# & Time (ms)	& Steps	& Search \\
	\midrule
	VC 0\_1	& 4426	& 5	& 0	\\
	VC 0\_2	& 5039	& 5	& 0	\\
	VC 0\_3	& 6324	& 5	& 0	\\
	\bottomrule
\end{tabular}
\end{figure}

The results of verification are summarized Figure \ref{fig:results}. As shown, the correctness of \texttt{Toggle} is straightforward enough for the prover to establish automatically, however \texttt{Set\_First\_And\_Last}'s VC 0\_5 failed to be dispatched.

\begin{figure}[!htb]
\begin{verbatim}
Goal:
Element_At((|L'|  - |L'|), L') = b

Given:
    ...
1. |L| = Strip_Length and
2. |L|  > 1 and
3. |L''| = |L| and
4. Element_At(0, L'') = b
5. |L'| = |L''| and
6. Element_At((|L''|  - 1), L') = b)
\end{verbatim}
\caption{A problematic VC}
\label{fig:problematicVC}
\end{figure}

A cursory examination of the unproven VC, shown in Figure \ref{fig:problematicVC}, reveals that prover was not able to employ the seemingly relevant \texttt{given \#4} since \texttt{L"}, could not be considered (rightfully so) the same as \texttt{L'} due to the changing boolean value. There are several means of correcting this: The first involves amending the \texttt{ensures} clause of operation \texttt{Set} (which is used in \texttt{Set First And Last}) to intuitively say that ``the content of \textit{outgoing} \texttt{L} is equivalent to the that of the incoming \texttt{L} with the exception of position \texttt{i}. The second route -- perhaps the less preferable of the two -- involves adding new theorems to \texttt{String\_Theory} which state this.

FIGURE OUT AN ELEGANT WAY TO CONCLUDE THIS SECTION...

\subsection{Facilities}
\label{sec:facilities}
The most straightforward means of utilizing our previously defined Led strip component and its verified enhancements is a via \textit{facility} module. Figure \# depicts a simple

%mention performance stuff in future work section.

\begin{figure*}
\centering
\includegraphics[scale=.55]{figs/ast_traversal.pdf}
\caption{A RESOLVE operation AST, the walk call sequence, and sample translation output.}
\end{figure*}
\label{fig:ast}

% IN THIS SECTION SHOW SOME OF OUR TRANSLATED CODE... AND WALK THROUGH IT THE SAME WAY YOU DID WITH THE CONCEPT IN RESOLVE. BUT BEFORE YOU DO THAT, SHOW THE PICTURE (The one thats already here showing high level relationships and update it!).
\section{Implementation}
Development of the tool presented here can be logically partitioned into three distinct phases: 
\begin{enumerate}
\item Arriving at a translation model (or, strategy) for an efficient, reliable C representation of RESOLVE.
\item Implementing reusable mechanisms for carrying out the C code generation process.
\item Creation of a memory manager capable of safely allocating and freeing dynamic memory required by the generated code.
\end{enumerate}
In this section we discuss each of these phases, concluding with a demo illustrating each working in tandem to realize the example presented in section \ref{sec:specifiying}.

\subsection{C Translation Model}
One of the primary challenges in translating from RESOLVE to C is finding a suitable C analog for each RESOLVE module and the constructs allowable in each. Indeed, since we are dealing with an environment where functional correctness is a primary concern, it is important that the code generated by our tool represents as closely as possible the original RESOLVE source. In an effort to make such considerations, at the highest level, the C code generated makes special considerations for facilities, concepts, and realizations.

The relationship between the RESOLVE module type and c code generated is depicted at a high level in Figure \#.

Unsurprisingly, concepts are represented in C by a single header .h listing the various methods

\begin{figure}
\centering
\includegraphics[scale=.55]{figs/relationship.pdf}
\caption{RESOLVE module types and their corresponding C representation}
\end{figure}
\label{fig:ast}

\subsection{Translator Implementation}

\subsubsection{AST Traversal}
Translation is performed over the course of a traversal of RESOLVE's abstract syntax tree (AST). The traversal mechanism used is a dervivative of the visitor pattern that provides a SAX-dom style pre-post traversal over all nodes in the tree. Thus, for any given node present, a total of two visits occur: One corresponding to the node being `hit' during the pre traversal stage, and one for the post.

@@

\begin{verbatim}
Operation Nothing(); Procedure
        Var X : Integer;
        X := 3;
end Nothing;
\end{verbatim}

Shown in Figure \ref{fig:ast} is the AST representation of operation \texttt{Nothing}. Here, language constructs are represented as labeled boxes, while the actual traversal over these constructs -- and the order in which it is performed -- is communicated on the right via the call stack. Each of these calls are received in the translator in the order in which they are visited within the tree. It is up to the client (in this case, the author of the C-translator) to decide which of these methods they wish to override and perform custom actions within.

We feel this particular traversal pattern lends itself to task of source to source translation for the following reasons:

\begin{itemize}
\item A visit method for a construct provides a convenient encapsulation of all the logic required to translate RESOLVE construct $x$ to appropriate C construct $y$.

@@

\item Finally, only the visit methods for the constructs we currently wish to process need to be overridden. This makes it easy to tweak and optimize the size of the translator's codebase, implementing only visit method for language constructs that explicitly require translation.  
\end{itemize}

\subsubsection{Translation output}
Output of translated code is done using \textit{Stringtemplate} -- a third-party tool written in Java that allows users to define parameterizable templates. Like the name suggests, a template is simply ``a document with holes" that the user choses when and how to fill.

An example C function definition template is shown below.

@@

User supplied attributes, enclosed in \texttt{<..>}, indicates the position of that attribute relative to others. It is entirely up to the user to define which attributes to fill in, and how complex they want them to be. For example, the user might choose to fill the \texttt{params} attribute with a simple string, or a separately defined \texttt{parameter} template, which in turn might use another separately defined \texttt{type} template.

@@

In the context of language translation, these templates, when stored on a stack and manipulated over the course of the aforementioned AST traversal, help simplify the task of producing complicated, structured blocks of C output. For instance, upon visiting \texttt{preOperationDec}, a \texttt{function\_def} template can be instantiated by the client and pushed onto a global translation stack with its \texttt{name}, return \texttt{type}, and \texttt{modifier} attributes filled in. As \texttt{preOperationDec}'s children are visited, the \texttt{function\_def} template currently at the top of the stack receives similarly constructed parameter, variable, and statement templates from the nodes being walked. Upon reaching \texttt{postOperationDec}, we can be assured that the function has been completely filled in with the appropriate templates -- assuming the user has implemented the children's visit methods.

Hence, the only actual work being performed within visit methods is forwarding appropriate information from tree-node it represents, to an externally defined template. This allows us to exploit (in shameless design pattern parlance) a strict model view controller (MVC) separation in the translator's codebase between the mechanism that does the AST visiting (controller), the tree nodes from which we're adding information to templates (model), and the external file containing all available C language templates (view).

@@

\subsection{Memory Allocation}

Dynamic memory allocation does not lend itself well with an embedded programming paradigm. With hardware constraints, embedded systems have limited capacities on RAM. The telos motes are limited to 128 bytes of RAM. Many programs, however, require dynamic memory allocation to be used, including RESOLVE, in order to make it extensible. In this section we introduce a stack memory allocator.

\subsubsection{Allocation using salloc}

The \texttt{salloc} is a first fit memory allocator that is provided memory of fixed size that is determined at compile time. It requires a small section of meta-data called a block which stores information of about the size of memory allocated, neighboring blocks, as well if the block is free or not. A sample representation of the stack as shown in figure \ref{fig:stack}, is a typical example of allocated memory on the stack.

\begin{figure}[!htb]
\centering
\includegraphics[scale=.55]{figs/stack.pdf}
\caption{Representation of memory stack}
\end{figure}
\label{fig:stack}

\subsubsection{Deallocation using sfree}

A memory allocator must provide a mechanism to release  or free memory in order to indicate that it is not being used and can be reallocated for something else. The \texttt{sfree} operation is an analogous implementation to the standard C \texttt{free} function for the stack.

\begin{figure}[!htb]
\centering
\includegraphics[scale=.55]{figs/free.pdf}
\caption{sfree to free memory}
\end{figure}
\label{fig:free}

\subsubsection{Optimizations}

A common problem that can occur in memory allocation is fragmentation.  As show in figure ??, fragmentation can lead to inefficient and increased memory usage. This problem is magnified on embedded systems with limited memory capabilities. Simple optimizations can be made however to reduce fragmentation. Splitting is one optimization that \texttt{salloc} uses to maximize memory usage. Shown in figure \ref{fig:split}, previously allocated blocks that have been freed, can be split to the size that is required by a call to \texttt{salloc}. Another means of optimizing memory usage is fusing together freed blocks. When a call to \texttt{sfree()} is made, neighboring blocks are coalesced together, as shown in figure \ref{fig:fuse}.

\begin{figure}[!htb]
\centering
\includegraphics[scale=.55]{figs/split.pdf}
\caption{block-splitting}
\end{figure}
\label{fig:split}

\begin{figure}[!htb]
\centering
\includegraphics[scale=.55]{figs/fuse.pdf}
\caption{fusing blocks}
\end{figure}
\label{fig:fuse}
%A stack-based approached to dynamic memory allocation requires a fixed size of memory to be set prior to compilation.

@@

Special thanks to Mike Kabanni, Mark Todd, and Bruce Weide whose suggestions and initial contributions in constructing the current model of RESOLVE to C translation made this work possible.
